Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation

### 2013-08-05 exp1/iswc2013.tex
Helen's feedback
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@637 43ef1190-90fd-4dd7-a769-d0121f02b80e

One important aspect of science is the structured approach to falsify
hypotheses, and thus is important that assertions about performance
are formulated as hypotheses that can be falsified.

Thus, attacking the complexity problem by eliminating parameters we
think influence the outcome (e.g. caching) for simplicity is untenable
as scientific practice. We must instead tackle this problem by
enhancing our capability of testing complex systems and to identify
situations where our assumptions are flawed. Such constraints can only
be made when we are highly confident that a certain parameter does not
have any influence.

@@

For example, one may measure the execution time of a SPARQL endpoint
by measuring it when it contains 1 or 2 million triples.  Then, the
execution time is the response variable, the factor is the number of
triples and it has two levels, 1 or 2 million triples. We have chosen
in this paper to only deal with the relatively straightforward aspects
of the DoE formalism, and therefore constrain the experiment strictly
to two levels.

@@

In many fields of science, experimental economy is extremely
important, and so, DoE offers extensive methodology to run a certain
fraction of the runs at the price of explanatory power. Such
experiments are called \emph{fractional factorial experiments} and is
covered in Section~\ref{sec:frac}.

@@

As the focus of this paper is didactical, the actual measurements are
much simpler than what has been used in benchmarking. We have used the
data-set of DBPedia SPARQL Benchmark\cite{mxro:Morsey2011DBpedia}, but
only consider the smallest of their data-sets (that we found to be
more than 15 MTriples). Moreover, we have taken subsets of that
data-set by using the last 1 or 2 MTriples from their file.

@@

A real optimization is beyond the scope of this paper. Instead we have
performed a simulation that enables us to understand the effect of the
changes.  In a real-world case, there is often the case that an
optimization has negative side-effects, and we would need to simulate
both the optimization and the side-effects. To do that, we degraded
the performance of the 4store SPARQL implementation by inserting
\texttt{sleep} statements. Specifically we inserted the C statement
\texttt{sleep(2)} on line~920 in \texttt{src/frontend/query.c} on
level~1, to simulate the optimization. This has the effect of delaying
execution for 2~seconds for every block for all kinds of joins. On
level 2, we inserted \texttt{usleep(2000)} on line 987 in
\texttt{src/frontend/filter.c} to simulate the negative
side-effect. This delays execution for 2 milliseconds every time the
\texttt{langMatches} SPARQL function is called.

@@

Finally note that there are two common issues in benchmarking we do
not consider: we do not allow a warm-up run, nor do we take into
account that the server may cache all or parts of the result set. The
reasons for this choice will be discussed later.

@@

We perform the test as a one-sided hypothesis test where $H_0$ is that
the mean of the vectors are equal and $H_1$ is that mean of the
vectors representing the new implementation is lower. In this example,
we find that the available data supports the assertion that the new
implementation is better with a high probability, $p=1.16 \cdot
10^{-07}$.

While it is interesting to test the overall hypothesis that the new
implementation is an improvement, it is also interesting to know if
there are cases that it fails. Again, we treat the experiment as a
$2^6$ experiment replicated 4 times. Now, for each of the 32 resulting
combinations of the environmental factors for each level of
``Implement'', we extract 4 values. We may now run a t-test for each
of the replications, as above. The results are tabulated in
Table~\ref{tab:pvaluesfull}. We see that there are significant
improvements in most cases, one may want to further investigate the
cases where they do not.

@@

The main objective of this paper is not to establish that the
optimization is significant, it is merely a simulation. It is to
establish a critical practice of evaluations. We must therefore
understand why some effects are significant. That
``Implement'' is significant is hardly a surprise, that is what we
worked for. That the main effects ``Union'', ``Optional'' and ``Lang''
are strong is also due to that these are fairly demanding things to
evaluate. Since the delay we inserted affects all kinds of joins, it
is also intuitive that interactions between ``Implement'' and those
that require joins are strong. The strong ``BGPComp:Lang'' and
``Lang:Union'' interactions might be due to that the many more triples
needs to be searched for a language tag in one case. However, the
positive ``Implement:TripleC'' interaction evades such explanations;
in fact, it hints that our optimization may not work as well for
larger databases. By carefully inspecting Table~\ref{tab:pvaluesfull},
we see that whenever ``TripleC'' is at the low level, the
new implementation is always best, only at the higher level, the old
implementation may be better. This is a clear indication that the
separation between levels for the sizes of the data-set is too small
and should be investigated further. Such reasoning should be applied
to all significant effects.

@@

We turn to the 64-run experiment and its normal plot in
Figure~\ref{fig:frac64normal}. We see that the we get more significant
effects, and that they correspond well to those we saw in the full
factorial experiment. They also all have an intuitive explanation as
above. However, we do not see the effects that we found worrisome in
the full factorial design. The ``Implement:TripleC'' interaction
emerges only for a level $\alpha=0.15$. To ensure that we discover
such cases will be a key challenge in further studies.

\begin{figure}[ht!]
  \centerline{%
    \includegraphics[width=0.96\textwidth]{frac64normal.pdf}}
  \caption{A normal plot of the Fractional Factorial Experiment with
    64 runs. The labelled points are considered significant by using
    the Lenth criterion. Triangles correspond to a level $\alpha=0.15$
    and crosses at $\alpha=0.05$. The use of different symbols have
    been added to FrF2 by the authors.}\label{fig:frac64normal}
\end{figure}

@@

The importance of finding flaws in the levels is highlighted by the
normal plot in Figure~\ref{fig:frac64hugenormal}. The smaller
experiment made it feasible to run with a larger data-set, this
experiment uses the original 15~MTriples data-set of
\cite{mxro:Morsey2011DBpedia}. The picture is now as hinted in the
previous discussion, for larger ``TripleC'', our simulated performance
is actually worse, as evident from the positive ``Implement''. We also
see that the interactions ``Implement:Lang'' and ``Implement:TripleC''
is highly significant. It is clear what happened: For larger
``TripleC'', the delay we introduced in the langMatches function has a
devastating effect, since it delays the execution for every triple
that the query engine applies the filter to, and thereby completely
dominates the execution time.

@@

Moreover, we note that the ``Machine'' factor has become highly
significant in the other end along with its ``Machine:TripleC''
interaction. One should take careful notice that the ``Machine''
factor encompasses many different things, that should be different
factors; CPU, disks, available RAM, installed software, tuning
parameters and so on. It is advantageous to have such broad factors in
the design in early experiments, because they may point out unexpected
effects that should be further scrutinized. The ``Machine'' factor is
a good example of usage that also helps us manage complexity: If
possible, one may group many factors into one if one is in doubt of
its relevance and break it up in subsequent experiments if deeper
understanding is needed. To some extent, ``BGPComp'' is also a broad
factor, since it encompasses many structures that are normally seen in
a Basic Graph Pattern. In this case, the cause of the performance
problem was highly audible: It was the old P-ATA disks of the larger
machine, but in general, such clues are not available to the
experimenter, therefore more factors are needed.

@@

We have argued that fixing certain parameters is untenable, but we
have allowed ourselves to disregard caching and warm-up runs. We are
aware that caching may be important in 4store, as a preliminary
experiment gave a response time of 53.9~s of a query with ``BGPComp''
at level~2 but other factors at level~1. When an OPTIONAL clause was
added, the response time dropped to 1.25~s. As this is not more
restrictive query, the most likely explanation is that the result of
the evaluation of the Basic Graph Pattern was cached, so only the
OPTIONAL left join was necessary to add in the second query.

We did this to illustrate \emph{randomization} as this is what that
makes it permissible, i.e. the order of execution is random, so the
benefit of caching is likely to apply randomly to different runs, for
example, had the run order been reversed in the previous example, it
could have been the query without OPTIONAL that would have
benefited. Unless the effect of caching is cumulative throughout the
experiment (which is possible), the randomization will have the effect
of reversing the runs randomly. The end result is that the effect of
so-called ``lurking variables'' such as caching, warm-up effects, etc,
contributes to total unexplained variance of the experiment, but
should not skew the results to invalidate the estimated effects.

@@

We saw in Section~\ref{sec:frac} that with a smaller experiment, fewer
interactions are significant, and so provides us with fewer clues to
assess the soundness of the experiment. This should be a focus of
further research.

Finally, we have only employed the simplest parts of the DoE theory,
what has been presented here is part of a much more general formalism
known as Orthogonal Arrays. The use of orthogonal arrays allow for
different numbers of levels, for much greater flexibility in total run
size, and for non-regular designs that can provide a good fit for the
complex problem of SPARQL endpoint evaluations.

@@

As evaluating SPARQL endpoints is inherently difficult a simplistic
experiment such as the one we designed should not hold up to
scrutiny. We set out to demonstrate how an experiment could be set up
using DoE, and show how it can be analyzed. In
sections~\ref{sec:full}~and~\ref{sec:frac}, we first saw how the
analysis correctly pointed out the most important effects, under
assumptions dictated by the factors and levels that were given. We saw
how the formalism provided a comprehensive view of the
experiment. Then, we saw that the formalism pointed out weaknesses
that could invalidate the experiment, and in
Section~\ref{sec:hugefrac} we saw that the experiment did indeed not
hold up to scrutiny.

@@

Also, experiments as small as the 32-run is not useful in estimating
any detailed characteristics and is thus of little use for science,
but could be useful in engineering: If the experiment has been
validated by a larger experiment, it could serve well in a Continuous
Integration system.

We have seen that we can perform a proper hypothesis test based on
summary statistics, and by using our expertise, we showed how to show
that the experiment was flawed, and we have cautioned that failure to
control unexplained variance may compromise our ability to do so. We
have also pointed out how this direction can provide more rigorous
evaluation practices.

>>>>

One important aspect of science is the structured approach to falsifying
hypotheses, and therefore is important that assertions about performance
are formulated as hypotheses that can be falsified.

Therefore, attacking the complexity problem by eliminating parameters we
think influence the outcome (e.g. caching) for simplicity is untenable
as scientific practice. We must instead tackle this problem by
enhancing our capability of testing complex systems and to identify
situations where our assumptions are flawed. Such constraints can only
be made when we are highly confident that a certain parameter does not
have any influence.

@@

For example, one may measure the execution time of a SPARQL endpoint
by measuring it when it contains 1 or 2 million triples.  Then, the
execution time is the response variable, the factor is the number of
triples and it has two levels, 1 or 2 million triples. We have chosen
in this paper to only deal with the relatively straightforward aspects
of the DoE formalism, and therefore constrained the experiment strictly
to two levels.

@@

In many fields of science, experimental economy is extremely
important, and so, DoE offers extensive methodology to run a certain
fraction of the runs at the price of explanatory power. Such
experiments are called \emph{fractional factorial experiments} and are
covered in Section~\ref{sec:frac}.

@@

As the focus of this paper is didactical, the actual measurements are
much simpler than those that have been used in benchmarking. We have used the
data-set of DBPedia SPARQL Benchmark\cite{mxro:Morsey2011DBpedia}, but
only consider the smallest of their data-sets (that we found to be
more than 15 MTriples). Moreover, we have taken subsets of that
data-set by using the last 1 or 2 MTriples from their file.

@@

A real optimization is beyond the scope of this paper. Instead we have
performed a simulation that enables us to understand the effect of the
changes.  In a real-world case, it is often the case that an
optimization has negative side-effects, and we would need to simulate
both the optimization and the side-effects. To do that, we degraded
the performance of the 4store SPARQL implementation by inserting
\texttt{sleep} statements. Specifically we inserted the C statement
\texttt{sleep(2)} on line~920 in \texttt{src/frontend/query.c} on
level~1, to simulate the optimization. This has the effect of delaying
execution for 2~seconds for every block for all kinds of joins. On
level 2, we inserted \texttt{usleep(2000)} on line 987 in
\texttt{src/frontend/filter.c} to simulate the negative
side-effect. This delays execution for 2 milliseconds every time the
\texttt{langMatches} SPARQL function is called.

@@

Finally, note that there are two common issues in benchmarking we do
not consider: we do not allow a warm-up run, nor do we take into
account that the server may cache all or parts of the result set. The
reasons for this choice will be discussed later.

@@

We perform the test as a one-sided hypothesis test where $H_0$ is that
the mean of the vectors are equal and $H_1$ is that mean sof the
vectors representing the new implementation is lower. In this example,
we find that the available data supports the assertion that the new
implementation is better with a high probability, $p=1.16 \cdot
10^{-07}$.

While it is interesting to test the overall hypothesis that the new
implementation is an improvement, it is also interesting to know if
there are cases where it fails. Again, we treat the experiment as a
$2^6$ experiment replicated 4 times. Now, for each of the 32 resulting
combinations of the environmental factors for each level of
``Implement'', we extract 4 values. We may now run a t-test for each
of the replications, as above. The results are tabulated in
Table~\ref{tab:pvaluesfull}. We see that there are significant
improvements in most cases; one may want to further investigate the
cases where they do not.

@@

The main objective of this paper is not to establish that the
optimization is significant, it is merely a simulation. It is to
establish a critical practice of evaluations. We must therefore
understand why some effects are significant. That
``Implement'' is significant is hardly a surprise, that is what we
worked for. That the main effects ``Union'', ``Optional'' and ``Lang''
are strong is also due to that these are fairly demanding things to
evaluate. Since the delay we inserted affects all kinds of joins, it
is also intuitive that interactions between ``Implement'' and those
that require joins are strong. The strong ``BGPComp:Lang'' and
``Lang:Union'' interactions might be due to that many more triples
need to be searched for a language tag in one of the levels of the
interacting factors. However, the
positive ``Implement:TripleC'' interaction evades such explanations;
in fact, it hints that our optimization may not work as well for
larger databases. By carefully inspecting Table~\ref{tab:pvaluesfull},
we see that whenever ``TripleC'' is at the low level, the
new implementation is always best. Only at the higher level, the old
implementation may be better. This is a clear indication that the
separation between levels for the sizes of the data-set is too small
and should be investigated further. Such reasoning should be applied
to all significant effects.

@@

We turn to the 64-run experiment and its normal plot in
Figure~\ref{fig:frac64normal}. We see that then we get more significant
effects, and that they correspond well to those we saw in the full
factorial experiment. They also all have an intuitive explanation as
above. However, we do not see the effects that we found worrisome in
the full factorial design. The ``Implement:TripleC'' interaction
emerges only for a level $\alpha=0.15$. To ensure that we discover
such cases will be a key challenge in further studies.

\begin{figure}[ht!]
  \centerline{%
    \includegraphics[width=0.96\textwidth]{frac64normal.pdf}}
  \caption{A normal plot of the Fractional Factorial Experiment with
    64 runs. The labelled points are considered significant by using
    the Lenth criterion. Triangles correspond to a level $\alpha=0.15$
    and crosses at $\alpha=0.05$. The use of different symbols has
    been added to FrF2 by the authors.}\label{fig:frac64normal}
\end{figure}

@@

The importance of finding flaws in the levels is highlighted by the
normal plot in Figure~\ref{fig:frac64hugenormal}. The smaller
experiment made it feasible to run with a larger data-set, this
experiment uses the original 15~MTriples data-set of
\cite{mxro:Morsey2011DBpedia}. The picture is now as hinted in the
previous discussion, for larger ``TripleC'', our simulated performance
is actually worse, as evident from the positive ``Implement''. We also
see that the interactions ``Implement:Lang'' and ``Implement:TripleC''
are highly significant. It is clear what happened: For larger
``TripleC'', the delay we introduced in the langMatches function has a
devastating effect, since it delays the execution for every triple
that the query engine applies the filter to, and thereby completely
dominates the execution time.

@@

Moreover, we note that the ``Machine'' factor has become highly
significant in the other end along with its ``Machine:TripleC''
interaction. One should take careful notice that the ``Machine''
factor encompasses many different things, that should be different
factors; CPU, disks, available RAM, installed software, tuning
parameters and so on. It is advantageous to have such broad factors in
the design in early experiments, because they may point out unexpected
effects that should be further scrutinized. The ``Machine'' factor is
a good example of usage that also helps us manage complexity: If
possible, one may group many factors into one if in doubt of
their relevance and break it up in subsequent experiments if deeper
understanding is needed. To some extent, ``BGPComp'' is also a broad
factor, since it encompasses many structures that are normally seen in
a Basic Graph Pattern. In this case, the cause of the performance
problem was highly audible: It was the old P-ATA disks of the larger
machine, but in general, such clues are not available to the
experimenter, therefore more factors are needed.

@@

We have argued that fixing certain parameters is untenable, but we
have allowed ourselves to disregard caching and warm-up runs. We are
aware that caching may be important in 4store, as a preliminary
experiment gave a response time of 53.9~s of a query with ``BGPComp''
at level~2 but other factors at level~1. When an OPTIONAL clause was
added, the response time dropped to 1.25~s. As this is not a more
restrictive query, the most likely explanation is that the result of
the evaluation of the Basic Graph Pattern was cached, so only the
OPTIONAL left join was necessary to add in the second query.

We did this to illustrate \emph{randomization} as this is what
makes it permissible, i.e. the order of execution is random, so the
benefit of caching is likely to apply randomly to different runs. For
example, had the run order been reversed in the previous example, it
could have been the query without OPTIONAL that would have
benefited. Unless the effect of caching is cumulative throughout the
experiment (which is possible), the randomization will have the effect
of reversing the runs randomly. The end result is that the effect of
so-called ``lurking variables'' such as caching, warm-up effects, etc,
contribute to the total unexplained variance of the experiment, but
should not skew the results to invalidate the estimated effects.

@@

We saw in Section~\ref{sec:frac} that with a smaller experiment, fewer
interactions are significant, and so provide us with fewer clues to
assess the soundness of the experiment. This should be a focus of
further research.

Finally, we have only employed the simplest parts of the DoE theory.
What has been presented here is part of a much more general formalism
known as Orthogonal Arrays. The use of orthogonal arrays allow for
different numbers of levels, for much greater flexibility in total run
size, and for non-regular designs that can provide a good fit for the
complex problem of SPARQL endpoint evaluations.

@@

As evaluating SPARQL endpoints is inherently difficult, a simplistic
experiment such as the one we designed should not hold up to
scrutiny. We set out to demonstrate how an experiment could be set up
using DoE, and show how it can be analyzed. In
sections~\ref{sec:full}~and~\ref{sec:frac}, we first saw how the
analysis correctly pointed out the most important effects, under
assumptions dictated by the factors and levels that were given. We saw
how the formalism provided a comprehensive view of the
experiment. Then, we saw that the formalism pointed out weaknesses
that could invalidate the experiment, and in
Section~\ref{sec:hugefrac} we saw that the experiment did indeed not
hold up to scrutiny.

@@

Also, experiments as small as the 32-run are not useful in estimating
any detailed characteristics and are thus of little use for science,
but could be useful in engineering: If the experiment has been
validated by a larger experiment, it could serve well in a Continuous
Integration system.

We have seen that we can perform a proper hypothesis test based on
summary statistics, and by using our expertise, we showed how to reveal
that the experiment was flawed, and we have cautioned that failure to
control unexplained variance may compromise our ability to do so. We
have also pointed out how this direction can provide more rigorous
evaluation practices.
